---
title: "Ground Motion Intensity Measure Evaluation Using Causdal Inference"
output: html_document
date: '2023-07-23'
author: "Henry Burton and Jack Baker"
---

# This code was developed by Henry Burton (UCLA) and Jack Baker (Stanford) to evaluate the effectiveness of ground motion intensity measures using causal infrerence. It is meant to accompany the following paper:

# Burton, H. V. and Baker, J. W. (2023) Evaluating the Effectiveness of Ground Motion Intensity Measures Through the Lens of Causal Inference. Earthquake Engineering Structural Dynamics (accepted for publication)

# Import libraries
```{r, include = FALSE}
# Import libraries
library(randomForest)
library(tictoc)
library(EnvStats)
```


# Define directory paths and import data
```{r}
# Define path to base directory
BaseDirectory = getwd()

# Define path to directory with raw data (.csv files)
InputDataDirectoryRaw = file.path(BaseDirectory,"Data","Raw")

# Define path to directory with causal inference results
ResultsDirectory = file.path(BaseDirectory,"Results","CausalInference")

# Import data
groundMotionSpectra <- read.csv(file.path(BaseDirectory,"Data","Raw","groundMotionSpectra.csv"),header = TRUE)
groundMotionParameters <- read.csv(file.path(BaseDirectory,"Data","Raw","groundMotionParameters.csv"),header = TRUE)
buildingInfo <- read.csv(file.path(BaseDirectory,"Data","Raw","BuildingInfo.csv"),header = TRUE)
PSDRResponses <- list(read.csv(file.path(BaseDirectory,"Data","Raw","1StoryPSDR.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","5StoryPSDR.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","9StoryPSDR.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","14StoryPSDR.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","19StoryPSDR.csv"),header = TRUE))
PFAResponses <- list(read.csv(file.path(BaseDirectory,"Data","Raw","1StoryPFA.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","5StoryPFA.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","9StoryPFA.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","14StoryPFA.csv"),header = TRUE),
                      read.csv(file.path(BaseDirectory,"Data","Raw","19StoryPFA.csv"),header = TRUE))
```

# Specify miscellaneous input variables
```{r}
# Number of buildings
NBuildings <- length(buildingInfo$numberOfStories)

# Number of IMs
NIMs <- 4

# Number of ground motions
Ngm <- 240

# Factor to convert pga from cm/s^2 to g
pgaFactor <- 0.00101971621

# Define factors associated with start and end periods for computing Saavg
TStartFactor <- 0.2
TEndFactor <- 3.0

# Random forest parameters
Mtry <- 2
NTrees <- 500

```

# Define function used to scale the IMs
```{r}
# Define function used to scale the IMs
scaleIMs <- function(PGA,PGV,SaT1,Saavg,NBuildings) {
  # Define number of GMs used
  noOfGMs <- length(PGA)
  
  
  # Initialize arrays used to store SaT1 and Saavg
  scaledSaT1 <- matrix(0,noOfGMs,NBuildings)
  scaledSaavg <- matrix(0,noOfGMs,NBuildings)
  
  # Scale PGA and PGV based on maximum values
  scaledPGA <- PGA/max(PGA)
  scaledPGV <- PGV/max(PGV)
    
  # Scale SaT1 abnd Saavg based on maximum values
  for (j in 1:NBuildings) {
    scaledSaT1[,j] <- SaT1[,j]/max(SaT1[,j])
    scaledSaavg[,j] <- Saavg[,j]/max(Saavg[,j])
  }
  # Create output array
  output <- list(scaledPGA,scaledPGV,scaledSaT1,scaledSaavg)
  return(output)  
}



```


# Define function used to extract IMs
```{r}
# Define function used to extract IMs
extractIMs <- function(mainIndexes,NBuildings,buildingInfo,groundMotionSpectra,groundMotionParameters,pgaFactor) {
  
  # Specify the number of ground motions
  noOfGMsUsed <- length(mainIndexes)
    
  # Initialize array used to store all SaT1 and Saavg values
  AllSaT1 <- matrix(0,Ngm,NBuildings)
  AllSaavg <- matrix(0,Ngm,NBuildings)
  
  # Extract unscaled IMs
  for (j in 1:NBuildings) {
    # Building period
    T1 <- buildingInfo$buildingPeriod[j]
    
    # Define start and end periods for computing Saavg
    TStart <- TStartFactor*T1
    TEnd <- TEndFactor*T1
      
    # Extract SaT1 values
    AllSaT1[,j] <- matrix(t(groundMotionSpectra[which(groundMotionSpectra$Period == T1),2:241]),Ngm,1)
    
    # Extract Sa values to compute Saavg
    SpectraForSaavg <- groundMotionSpectra[which.min(abs(groundMotionSpectra$Period - TStart)):which.min(abs(groundMotionSpectra$Period - TEnd)),2:(Ngm + 1)]
    
    
    # Compute Saavg
    for (i in 1:Ngm) AllSaavg[i,j] <- geoMean(SpectraForSaavg[,i])
  }
  
  # Extract pga and convert to units of "g"
  PGA <- as.matrix(pgaFactor*groundMotionParameters$PGA)

  # Extract PGV
  PGV <- as.matrix(groundMotionParameters$PGV)
  
  # Extract auxiliary IMs
  pgaAux <- PGA[-mainIndexes]
  pgvAux <- PGV[-mainIndexes]
  saT1Aux <- AllSaT1[-mainIndexes,]
  saavgAux <- AllSaavg[-mainIndexes,]
  
  # Extract main IMs
  pgaMain <- PGA[mainIndexes]
  pgvMain <- PGV[mainIndexes]
  saT1Main <- AllSaT1[mainIndexes,]
  saavgMain <- AllSaavg[mainIndexes,]

  # Create output array
  output <- list(pgaAux,pgvAux,saT1Aux,saavgAux,pgaMain,pgvMain,saT1Main,saavgMain)
  return(output)  
}


```



# Normalize the responses by the maximum over all considered ground motions
```{r}
#Initialize lists used to store normalized EDPs 
normalizedPSDRs <- vector(mode='list', length=NBuildings)
normalizedPFAs <- vector(mode='list', length=NBuildings)

# Loop over the number of buildings
for (j in 1:NBuildings) {
  
  # Extract the number of stories for the current building
  NStories <- buildingInfo$numberOfStories[j]

  # Extract PSDR Response and PFA Response for current building
  currentPSDRs <- PSDRResponses[[j]]
  currentPFAs <- PFAResponses[[j]]
  
  # Initialize matrix used to store normalized EDPs for current building
  currentNormalizedPSDRs <- matrix(0,NStories,Ngm)
  currentNormalizedPFAs <- matrix(0,NStories + 1,Ngm)
  
  # Loop over the number of stories
  for (i in 1:NStories) {
    # Extract normalized peak responses
    currentNormalizedPSDRs[i,] <- as.matrix(currentPSDRs[i,]/max(currentPSDRs[i,]),1,Ngm)
    currentNormalizedPFAs[i,] <- as.matrix(currentPFAs[i,]/max(currentPFAs[i,]),1,Ngm)
  }
  
  # Add normalized EDPs for current building to list
  normalizedPSDRs[[j]] <- currentNormalizedPSDRs
  normalizedPFAs[[j]] <- currentNormalizedPFAs
}

```
# Extract maximum of the normalized peak responses
```{r}
# Initialize arrays used to store maxPSDRResponse and maxPFAResponse responses
maxPSDRResponse <- matrix(0,Ngm,NBuildings)
maxPFAResponse <- matrix(0,Ngm,NBuildings)

# Loop over the number of buildings
for (j in 1:NBuildings) {

  # Extract PSDR Response and PFA Response for current building
  currentPSDRs <- normalizedPSDRs[[j]]
  currentPFAs <- normalizedPFAs[[j]]
  
  # Loop over the number of ground motions
  for (i in 1:Ngm) {
    # Extract peak responses
    maxPSDRResponse[i,j] <- max(currentPSDRs[,i])
    maxPFAResponse[i,j] <- max(currentPFAs[,i])
  }
}

```

# Perform causal inference considering only maxPSDR and maxPFA
```{r}
# Define indexes used for main set 
mainIndexes <- matrix(sample(1:Ngm, Ngm/2, replace = F),Ngm/2,1)
mainIndexes <- as.matrix(mainIndexes)

# Call function used to extract main and auxiliary IM sets
output <- extractIMs(mainIndexes,NBuildings,buildingInfo,groundMotionSpectra,groundMotionParameters,pgaFactor)

# Define auxiliary IM set
pgaAux <- output[[1]]
pgvAux <- output[[2]]
saT1Aux <- output[[3]]
saavgAux <- output[[4]]

# Define main IM set
pgaMain <- output[[5]]
pgvMain <- output[[6]]
saT1Main <- output[[7]]
saavgMain <- output[[8]]

# Scale auxiliary IM set
output <- scaleIMs(pgaAux,pgvAux,saT1Aux,saavgAux,NBuildings)
pgaAux <- output[[1]]
pgvAux <- output[[2]]
saT1Aux <- output[[3]]
saavgAux <- output[[4]]

# Scale main IM set
output <- scaleIMs(pgaMain,pgvMain,saT1Main,saavgMain,NBuildings)
pgaMain <- output[[1]]
pgvMain <- output[[2]]
saT1Main <- output[[3]]
saavgMain <- output[[4]]

# Create main and auxiliary EDP sets
psdrAux <- maxPSDRResponse[-mainIndexes,]
pfaAux <- maxPFAResponse[-mainIndexes,]
psdrMain <- maxPSDRResponse[mainIndexes,]
pfaMain <- maxPFAResponse[mainIndexes,]

# Create auxiliary causal parameters
mAux <- groundMotionParameters$Magnitude[-mainIndexes]
rAux <- groundMotionParameters$BooreJoynerDistance[-mainIndexes]
epAux <- groundMotionParameters$epsilon1[-mainIndexes]
vs30Aux <- groundMotionParameters$Vs30[-mainIndexes]

# Create main causal parameters
mMain <- groundMotionParameters$Magnitude[mainIndexes]
rMain <- groundMotionParameters$BooreJoynerDistance[mainIndexes]
epMain <- groundMotionParameters$epsilon1[mainIndexes]
vs30Main <- groundMotionParameters$Vs30[mainIndexes]
causalMain <- cbind(mMain,rMain,epMain,vs30Main)

# Define array used to store maxPSDR and maxPFA causal effect
causalEffectMaxPSDR <- matrix(0,NBuildings,NIMs)
causalEffectMaxPFA <- matrix(0,NBuildings,NIMs)

# Loop over the number of buildings
for (j in 1:NBuildings) {
  # Extract the SaT1 and Saavg for the current building
  currentSaavgAux <- saavgAux[,j]
  currentSaT1Aux <- saT1Aux[,j]
  currentSaavgMain <- saavgMain[,j]
  currentSaT1Main <- saT1Main[,j]
  
  # Assemble IMAux and IMMain
  IMAux <- cbind(pgaAux,pgvAux,currentSaT1Aux,currentSaavgAux)
  IMMain <- cbind(pgaMain,pgvMain,currentSaT1Main,currentSaavgMain)
  
  # Loop over the number of IMs
  for (i in 1:NIMs) {
    # Using the auxiliary data, construct RF model with M and R as features and IM as the response variable i.e., g(X)
    currentIMRFModel <- randomForest(IMAux[,i] ~ mAux + 
                                       rAux, mtry = Mtry,
                                       na.action = na.omit,ntree = NTrees)
  
    # Use current RF model to predict the main data IM
    predictedCurrentIM <- predict(currentIMRFModel, newdata = causalMain)
          
    # Using the auxiliary data, construct RF model with M and R as features and EDP as the response variable i.e., f(X)
      #print(psdrAux[,j])
    PSDRRFModel <- randomForest(psdrAux[,j] ~ mAux + 
                                       rAux, mtry = Mtry,
                                       na.action = na.omit,ntree = NTrees)
    PFARFModel <- randomForest(pfaAux[,j] ~ mAux + rAux,
                                       mtry = Mtry,na.action = na.omit,ntree = NTrees)
      
    # Use RF model to main data EDPs
    predictedPSDR <- predict(PSDRRFModel, newdata = causalMain)
    predictedPFA <- predict(PFARFModel, newdata = causalMain)
        
    # Compute the IM residuals
    currentIMResiduals <- predictedCurrentIM - IMMain[,i]
    
    # Compute the EDP residuals
    PSDRResiduals <- predictedPSDR - psdrMain[,j]
    PFAResiduals <- predictedPFA - pfaMain[,j]
        
    # Regress EDP residuals against IM residuals
    PSDRIMModel <- lm(PSDRResiduals ~ currentIMResiduals)
    PFAIMModel <- lm(PFAResiduals ~ currentIMResiduals)
        
    # Compute the causal effect (CE)
    causalEffectMaxPSDR[j,i] <- PSDRIMModel$coefficients[2]
    causalEffectMaxPFA[j,i] <- PFAIMModel$coefficients[2]

  }
}

# Print results
print(causalEffectMaxPSDR)
print(causalEffectMaxPFA)

# Save results
write.table(causalEffectMaxPSDR, file.path(BaseDirectory,"Results","CausalInference","causalEffectMaxPSDR.txt"), qu = F, row = F,col = F)
write.table(causalEffectMaxPSDR, file.path(BaseDirectory,"Results","CausalInference","causalEffectMaxPFA.txt"), qu = F, row = F,col = F)




```

# Perform causal inference considering full profile response i.e., PSDR and PFA
```{r}
# Initialize list used to store causal effects
causalEffectPSDR <- list()
causalEffectPFA <- list()

# Loop over the number of buildings
for (j in 1:NBuildings) {
  # Number of stories in current building
  NStories <- buildingInfo$numberOfStories[j]
  
  # Define array used to store PSDR and PFA causal effect for current building
  currentCausalEffectPSDR <- matrix(0,NStories,NIMs)
  currentCausalEffectPFA <- matrix(0,NStories,NIMs)
  currentR2FromCausalModelPSDR <- matrix(0,NStories,NIMs)
  currentR2FromCausalModelPFA <- matrix(0,NStories,NIMs)
  
  # Loop over the number of stories
  for (k in 1:NStories) {
    # Extract the SaT1 and Saavg for the current building
    currentSaavgAux <- saavgAux[,j]
    currentSaT1Aux <- saT1Aux[,j]
    currentSaavgMain <- saavgMain[,j]
    currentSaT1Main <- saT1Main[,j]
    
    # Assemble IMAux and IMMain
    IMAux <- cbind(pgaAux,pgvAux,currentSaT1Aux,currentSaavgAux)
    IMMain <- cbind(pgaMain,pgvMain,currentSaT1Main,currentSaavgMain)
    
    # Extract PSDR and PFA for current building/story/floor
    currentPSDR <- t(as.matrix(normalizedPSDRs[[j]][k,]))
    currentPFA <- t(as.matrix(normalizedPFAs[[j]][k,]))
    
    # Create Main and Auxiliary EDP Sets
    psdrAux <- currentPSDR[-mainIndexes]
    pfaAux <- currentPFA[-mainIndexes]
    psdrMain <- currentPSDR[mainIndexes]
    pfaMain <- currentPFA[mainIndexes]
    
    # Loop over the number of IMs
    for (i in 1:NIMs) {
      # Using the auxiliary data, construct RF model with M and R as features and IM as the response variable i.e., g(X)
      currentIMRFModel <- randomForest(IMAux[,i] ~ mAux + 
                                       rAux, mtry = Mtry,
                                       na.action = na.omit,ntree = NTrees)
  
      # Use current RF model to predict the main data IM
      predictedCurrentIM <- predict(currentIMRFModel, newdata = causalMain)
          
      # Using the auxiliary data, construct RF model with  M and R as features and EDP as the response variable i.e., f(X)
      PSDRRFModel <- randomForest(psdrAux ~ mAux + 
                                       rAux, mtry = Mtry,
                                       na.action = na.omit,ntree = NTrees)
      PFARFModel <- randomForest(pfaAux ~ mAux + rAux,
                                       mtry = Mtry,na.action = na.omit,ntree = NTrees)
        
      # Use RF model to main data EDPs
      predictedPSDR <- predict(PSDRRFModel, newdata = causalMain)
      predictedPFA <- predict(PFARFModel, newdata = causalMain)
          
      # Compute the IM residuals
      currentIMResiduals <- predictedCurrentIM - IMMain[,i]
      
      # Compute the EDP residuals
      PSDRResiduals <- predictedPSDR - psdrMain
      PFAResiduals <- predictedPFA - pfaMain
          
      # Regress EDP residuals against IM residuals
      PSDRIMModel <- lm(PSDRResiduals ~ currentIMResiduals)
      PFAIMModel <- lm(PFAResiduals ~ currentIMResiduals)
          
      # Compute the causal effect (CE)
      currentCausalEffectPSDR[k,i] <- PSDRIMModel$coefficients[[2]]
      currentCausalEffectPFA[k,i] <- PFAIMModel$coefficients[[2]]
      
      # Extract the R^2 value
      currentR2FromCausalModelPSDR[k,i] <- summary(PSDRIMModel)$r.squared
      currentR2FromCausalModelPFA[k,i] <- summary(PFAIMModel)$r.squared
      
    }
  }
  # Append to list of causal effects
  causalEffectPSDR[[j]] <- currentCausalEffectPSDR
  causalEffectPFA[[j]] <- currentCausalEffectPFA
  
  # Write to file
  filepath <- paste(file.path(BaseDirectory,"Results","CausalInference"),"/causalEffectPSDR",NStories,"Story.txt",sep = "", collapse = NULL)
  write.table(currentCausalEffectPSDR,filepath, qu = F, row = F,col = F)
  filepath <- paste(file.path(BaseDirectory,"Results","CausalInference"),"/causalEffectPFA",NStories,"Story.txt",sep = "", collapse = NULL)
  write.table(currentCausalEffectPFA,filepath, qu = F, row = F,col = F)
  
}
```